---
title: "Cotton Incorporated funded project - Quarterly Report"
format: docx
author: Leonardo Bastos, University of Georgia
---
```{r setup}
#| include: false
library(sf)
library(tidyverse)

theme_set(theme_gray()+
            theme(panel.grid.minor = element_blank())
            )
```

```{r df}
df <- read.csv("../data/daymet/fieldweather_final.csv") %>%
  mutate(region = case_when(
    grepl("CA|AZ|NM|NV", loc) ~ "FarWest",
    grepl("TX|OK|KS", loc) ~ "Southwest",
    grepl("MO|AR|LA|MS|TN", loc) ~ "Midsouth",
    grepl("AL|GA|SC|NC|VA|FL", loc)  ~ "Southeast",
    TRUE ~ "Other"
  ))

df
```


This quarterly report describes progress towards the final goal of utilizing the USDA long-term cotton variety trial and publicly available weather data to understand the historical effects of weather on cotton fiber quality.  

# Achieved tasks in the last quarter  
During this quarter, the following was achieved:  
- Downloaded, processed, and merged USDA cotton variety trial data sets into one data frame.  

- Appended each study site with the geographical coordinates of the centroid of nearest town.  

- Used site centroids to retrieve historical weather data from [Dayment](https://daymet.ornl.gov):  
  - Temporal resolution: daily  
  - Temporal extent: from 1980 through 2020
  - Spatial resolution: 1 x 1 km  
  - Primary weather variables:  
    - Precipitation (mm)  
    - Solar radiation (W/m2)  
    - Minimum air temperature (C)  
    - Maximum air temperature (C)  
    - Vapor pressure deficit (Pa)  

- Used primary weather variables to derive secondary weather variables:  
  - Temperature amplitude (C)  
  - Growing degree days  

- Summarized primary and secondary weather variables by summing (precipitation, solar radiation, growing degree days) or averaging (all others) across different temporal scales including:  
  - Growing season (April through November)  
  - Every month   
  - Every 2 months  
  - Every 4 months  

- Extracted the best linear unbiased estimators (BLUEs) of one fiber quality parameter (strength) across all site-years in the data set.  

- Merged summarized weather variables with fiber strength BLUEs in the same data set for machine learning training:  
  - Number of unique sites: 69  
  - Number of years: 41    
  - Number of site-years: 1,103  
  - Number of summarized weather predictor variables: 151    

# Future tasks  
- Train random forest models in the context of machine learning (data split, hyperparameter optimization, predictive power assessment) to predict cotton fiber strength as a function of weather. A different model will be trained for each of the summarized temporal scales (from monthly to entire growing season), and one including all temporal scales in same model.  

- Once the workflow has been developed for cotton fiber strength, similar analysis will be conducted for the remaining fiber quality variables (length, uniformity, micronaire).  



