---
title: "Cotton Incorporated funded project - Quarterly Report"
format: docx
author: Assistant Professor, University of Georgia
subtitle: Leonardo Bastos
---
```{r setup}
#| include: false
library(sf)
library(tidyverse)

theme_set(theme_gray()+
            theme(panel.grid = element_blank())
            )
```

```{r df}
#| echo: false
df <- read.csv("../data/daymet/fieldweather_final.csv") %>%
  mutate(region = case_when(
    grepl("CA|AZ|NM|NV", loc) ~ "FarWest",
    grepl("TX|OK|KS", loc) ~ "Southwest",
    grepl("MO|AR|LA|MS|TN", loc) ~ "Midsouth",
    grepl("AL|GA|SC|NC|VA|FL", loc)  ~ "Southeast",
    TRUE ~ "Other"
  ))

```


This quarterly report describes progress towards the final goal of utilizing the USDA long-term cotton variety trial and publicly available weather data to understand the historical effects of weather on cotton fiber quality.  

# Achieved tasks in the last quarter  
During this quarter, the following was achieved:  

- Downloaded, processed, and merged USDA cotton variety trial data sets into one data frame.  

- Appended each study site with the geographical coordinates of the centroid of nearest town.  

- Used site centroid to retrieve historical weather data from [Dayment](https://daymet.ornl.gov):  

  - Temporal resolution: daily
  - Temporal extent: from 1980 through 2020
  - Spatial resolution: 1 x 1 km
  - Primary weather variables:  
    - Day length (h)
    - Precipitation (mm)
    - Solar radiation (W/m2)
    - Minimum air temperature (C)
    - Maximum air temperature (C)
    - Vapor pressure deficit (Pa)


```{r}
#| fig-cap: Figure 1. Distribution of daily primary weather variables (summarized by the growing season). dayl is day length (h), prcp is precipitation (mm), srad is solar radiation (W/m2), tmax is maximum air temperature (C), tmin is minimum air temperature (C), vp is vapor pressure deficit. 
#| echo: false
#| fig-width: 7
df %>%
  dplyr::select(contains("growingseason")) %>%
  dplyr::select(!contains(c("gdd", "tamp"))) %>%
  mutate(dayl_mean_growingseason = dayl_mean_growingseason/3600) %>%
  pivot_longer(cols = dplyr::everything()) %>%
  mutate(name = str_remove(name, "_growingseason")) %>%
  ggplot(aes(x=value))+
  geom_density(aes(fill=name), alpha = .7)+
  facet_wrap(~name, scales = "free")+
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title = element_blank()
        )
```


- Used primary weather variables to derive secondary weather variables:
  - Temperature amplitude (C)
  - Growing degree days
  
```{r}
#| fig-cap: Figure 2. Distribution of daily secondary weather variables (summarized by the growing season). tamp is temperature amplitude, gdd is growing degree days.   
#| echo: false
#| fig-height: 2

df %>%
  dplyr::select(contains("growingseason")) %>%
  dplyr::select(contains(c("gdd", "tamp"))) %>%
  pivot_longer(cols = dplyr::everything()) %>%
  mutate(name = str_remove(name, "_growingseason")) %>%
  ggplot(aes(x=value))+
  geom_density(aes(fill=name), alpha = .7)+
  facet_wrap(~name, scales = "free")+
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title = element_blank()
        )
```


- Summarized primary and secondary weather variables by summing (precipitation, solar radiation, growing degree days) or averaging (all others) across different temporal scales including:  
  - Every month
  - Every 2 months
  - Every 4 months
  - Growing season (April through November)  
  
```{r}
#| fig-cap: Figure 3. Distribution of minimum air temperature summarized at different temporal scales. 
#| echo: false    
#| fig-width: 7
df %>%
  dplyr::select(contains("tmin")) %>%
  pivot_longer(cols = dplyr::everything()) %>%
    mutate(name = str_remove(name, "_mean")) %>%
  mutate(name = factor(name,
                       levels = c("tmin_April",
                                  "tmin_May",
                                  "tmin_June",
                                  "tmin_July",
                                  "tmin_August",
                                  "tmin_September",
                                  "tmin_October",
                                  "tmin_November",
                                  
                                  "tmin_April_May",
                                  "tmin_June_July",
                                  "tmin_August_September",
                                  "tmin_October_November",
                                  
                                  "tmin_April_May_June_July",
                                  "tmin_August_September_October_November",
                         
                         
                         
                         "tmin_growingseason"
                                  
                                  )
                       
                       )) %>%
  ggplot(aes(x=value))+
  geom_density(aes(fill=name), alpha = .7)+
  facet_wrap(~name, scales = "free_y")+
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title = element_blank()
        )

```


- Extracted the best linear unbiased estimators (BLUEs) of one fiber quality parameter (strength) across all site-years in the data set.  

```{r}
#| fig-cap: Figure 3. Distribution of cotton fiber strength best linear unbiased estimator.   
#| echo: false
#| fig-height: 3
#| warning: false
df %>%
  ggplot(aes(x=str_emmean))+
  geom_density(fill = "blue", alpha = .7)+
  labs(x="Strength best linear unbiased estimator")+
  theme(legend.position = "none",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()
        )
```
```{r}
#| fig-cap: Figure 4. Cotton fiber strength time-series.
#| echo: false
#| warning: false
df %>%
  ggplot(aes(x=year, y=str_emmean))+
  geom_point(shape = 21, size = 2, alpha = .5)+
  labs(y="Strength best linear unbiased estimator",
       x="Year")+
  geom_smooth(method = "lm", color = "blue")

```


- Merged summarized weather variables with fiber strength BLUEs in the same data set for machine learning training:  
  - Number of unique sites: 69
  - Number of years: 41
  - Number of site-years: 1,103
  - Number of summarized weather predictor variables: 151

# Future tasks  
- Train random forest models in a machine learning framework (data split, hyperparameter optimization, predictive power assessment) to predict cotton fiber strength as a function of weather. A different model will be trained for each of the summarized temporal scales (from monthly to entire growing season), and one including all temporal scales in same model.  

- Once the workflow has been developed for cotton fiber strength, similar analysis will be conducted for the remaining fiber quality variables (length, uniformity, micronaire).
